|

### 2024-07-10 doc/paper_mine.qmd
Prepared first successful draft of paper.pdf

---
title: |
  | Assignment II: 
  | Adopting an Open Science Workflow \vspace{1cm}
author: |
  | Melisa Mazaeva 
  | Accounting Reading Group – Summer 2024
  | Humboldt-Universität zu Berlin 
  | melisa.mazaeva@student.hu-berlin.de
date: today
date-format: MMM D, YYYY [\vspace{1cm}]
abstract: |
  | This project uses the TRR 266 Template for Reproducible Empirical Accounting Research (TREAT) and provides an infrastructure for open science oriented empirical projects. It has been adapted to analyze residual income based on grouped P/B ratios of U.S. public firms using external Worldscope data sets. It showcases a reproducible workflow integrating Python scripts and data analysis. The analysis is performed using a Python script that processes input data files from Wolrdscope and generates output documentation files with the results. This code base, adapted from TREAT, should give you an overview on how the template is supposed to be used for my specific project and how to structure a reproducible empirical project.
  | \vspace{6cm}
bibliography: references.bib
biblio-style: apsr
format:
  pdf:
    documentclass: article
    number-sections: true
    toc: false
fig_caption: yes
fontsize: 11pt
ident: yes
always_allow_html: yes
header-includes:
  - \usepackage[nolists]{endfloat}    
  - \usepackage{setspace}\doublespacing
  - \setlength{\parindent}{4em}
  - \setlength{\parskip}{0em}
  - \usepackage[hang,flushmargin]{footmisc}
  - \usepackage{caption} 
  - \captionsetup[table]{skip=24pt,font=bf}
  - \usepackage{array}
  - \usepackage{threeparttable}
  - \usepackage{adjustbox}
  - \usepackage{graphicx}
  - \usepackage{csquotes}
  - \usepackage[margin=1in]{geometry}
---

@@

The primary objective is to demonstrate a reproducible and collaborative research workflow by calculating and analyzing residual incomes and Price-to-Book (P/B) ratios. The analysis includes filtering relevant data, identifying key financial years, grouping firms based on their P/B ratios, and calculating delfated residual incomes for each firm/year.
Assignment I aims to replicate an empirical table that involves calculating the residual income of US firms over a defined period and examining the relationship between residual income and the Price-to-Book (P/B) ratio. The replication process includes data preparation, cleaning, normalization, and the application of statistical methods to compute and interpret financial metrics using Python. By following a step-by-step approach, I provide insights into the trends of residual earnings across different P/B groups, identical to trends presented by Penman (2013).
This document presents the analysis performed in Assignment 1, adapted to the open science workflow. Note that intermediate check steps have been omitted.

>>>>

---
title: |
  | Assignment II: 
  | Adopting an Open Science Workflow \vspace{1cm}
author: |
  | Melisa Mazaeva 
  | Accounting Reading Group – Summer 2024
  | Humboldt-Universität zu Berlin 
  | melisa.mazaeva@student.hu-berlin.de
date: today
date-format: MMM D, YYYY [\vspace{1cm}]
abstract: |
  | This project uses the TRR 266 Template for Reproducible Empirical Accounting Research (TREAT) and provides an infrastructure for open science oriented empirical projects. It has been adapted to analyze residual income based on grouped P/B ratios of U.S. public firms using external Worldscope data sets. It showcases a reproducible workflow integrating Python scripts and data analysis. The analysis is performed using a Python script that processes input data files from Wolrdscope and generates output documentation files with the results. This code base, adapted from TREAT, should give you an overview on how the template is supposed to be used for my specific project and how to structure a reproducible empirical project.
  | \vspace{6cm}
bibliography: references.bib
biblio-style: apsr
format:
  pdf:
    documentclass: article
    number-sections: true
    toc: false
fig_caption: yes
fontsize: 11pt
ident: yes
always_allow_html: yes
header-includes:
  - \usepackage[nolists]{endfloat}    
  - \usepackage{setspace}\doublespacing
  - \setlength{\parindent}{4em}
  - \setlength{\parskip}{0em}
  - \usepackage[hang,flushmargin]{footmisc}
  - \usepackage{caption} 
  - \captionsetup[table]{skip=24pt,font=bf}
  - \usepackage{array}
  - \usepackage{threeparttable}
  - \usepackage{adjustbox}
  - \usepackage{graphicx}
  - \usepackage{csquotes}
  - \usepackage{indentfirst}  # Added this line to ensure the first paragraph is indented
  - \usepackage[margin=1in]{geometry}
---

# List of Abbreviations
\setlength{\parindent}{0em}

**AMEX**: American Stock Exchange  
**CAPM**: Capital Asset Pricing Model  
**BVE**: Book Value of Equity  
**IDE**: Integrated Development Environment  
**MVE**: Market Value of Equity  
**NYSE**: New York Stock Exchange  
**P/B**: Price-to-Book  
**UK**: United Kingdom  
**US**: United States

\setlength{\parindent}{4em}

@@

The primary objective is to demonstrate a reproducible and collaborative research workflow by calculating and analyzing residual incomes and Price-to-Book (P/B) ratios. The analysis includes filtering relevant data, identifying key financial years, grouping firms based on their P/B ratios, and calculating delfated residual incomes for each firm/year.

Assignment I aims to replicate an empirical table that involves calculating the residual income of US firms over a defined period and examining the relationship between residual income and the Price-to-Book (P/B) ratio. The replication process includes data preparation, cleaning, normalization, and the application of statistical methods to compute and interpret financial metrics using Python. By following a step-by-step approach, I provide insights into the trends of residual earnings across different P/B groups, identical to trends presented by @Penman_2013. This document presents the analysis performed in Assignment I, adapted to the open science workflow. Note that intermediate check steps have been omitted.

# Assumptions and Notes

The aim of Assignment I is to replicate a specific empirical table that involves calculating the residual income of US firms over a defined period and examining the relationship between residual income and the Price-to-Book (P/B) ratio. The replication process involves data preparation, cleaning, and normalization, followed by the application of statistical methods to compute and interpret financial metrics. For Assignment I, I used the Python programming language to carry out the empirical analysis. Visual Studio Code was used as the Integrated Development Environment (IDE) for writing, debugging, and optimizing the Python code.

## Assumptions
1. A constant cost of equity is assumed to simplify the calculation of residual income across multiple years, making the analysis more straightforward and easier to interpret. This approach allows for a direct comparison of performance across firms and time periods without the increased complexity due to fluctuating rates. In contrary, estimating a dynamic cost of equity typically involves financial models such as the Capital Asset Pricing Model (CAPM), which would require additional data like risk-free rates and firm beta values that are not available.
2. The level of cost of equity capital is assumed to be 8% (re = 0.08), a reasonable average estimate for the period 1996-2015 in the US (@Goedhart_Koller_Williams_2002; @Damodaran_2024). This is assumption reflects the return equity investors typically expect given the level of risk during that period.
3. The entire Worldscope panel data provides the Book Value of Equity (BVE), Market Value of Equity (MVE), and net income as values for the end of the respective fiscal year.
4. The beginning of Year 0 data is equivalent to the end of the previous year because the financial data recorded at the end of one year represents the starting values for the following year. Therefore, residual income is deflated by the BVE at the end of the year before Year 0 to ensure comparability across different firms and time periods.
5. It is assumed that Penman (2013) restricted the possible P/B values to a range of 0 to 7 to exclude outliers. Hence, extreme P/B values that are negative and very high (greater than 7) are excluded to focus the analysis on firms with more stable and reasonable valuations, reducing the impact of outliers.

By following the steps provided in Section 4 and adhering to the assumptions made, I successfully replicated the analysis and produced the required table. A thorough step-by-step approach helped to understand and verify the outputs.

# Replication Steps

```{python}
#| label: step1
#| echo: true
#| output: true

import pandas as pd
import numpy as np
import os

# Paths to data files
static_data_path = '/Users/melissam/Documents/GitHub/arg_trr266/data/external/wscp_static.txt' 
panel_data_path = '/Users/melissam/Documents/GitHub/arg_trr266/data/external/wscp_panel.xlsx'

# Check if files exist
if not os.path.exists(static_data_path):
    raise FileNotFoundError(f"{static_data_path} does not exist.")
if not os.path.exists(panel_data_path):
    raise FileNotFoundError(f"{panel_data_path} does not exist.")

# Load the data from Worldscope files into pandas DataFrames
static_data = pd.read_csv(static_data_path, delimiter='\t')  # use '\t' due to tab-separation in the static file
panel_data = pd.read_excel(panel_data_path)

# Normalize column names to avoid issues caused by variation in column naming conventions
static_data.columns = static_data.columns.str.strip().str.lower().str.replace(' ', '_')
panel_data.columns = panel_data.columns.str.strip().str.lower().str.replace(' ', '_')

# Merge data on 'isin' to include country information in panel_data
merged_data = panel_data.merge(static_data[['isin', 'country']], on='isin', how='left')

# Print the merged data to check. Now we have added a column on countries in panel data
print("Step 1: Merged data sample on 'isin' from static and panel data:\n", merged_data.head(11))
print()  # add this to get a blank line for better readability
# To understand how many rows there are in the merged data
num_rows1 = len(merged_data)
print("Number of rows in merged data sample on 'isin' from static and panel data:", num_rows1)
print()
print()
```

```{python}
#| label: step2
#| echo: true
#| output: true

# Filter the merged data for US firms and years 1996-2015
us_firms = merged_data[merged_data['country'] == 'UNITED STATES']  # create a new DataFrame that contains only US firms
filtered_data = us_firms[(us_firms['year_'] >= 1996) & (us_firms['year_'] <= 2015)].copy()  # further filter the us_firms DataFrame to keep the task's required year range

# Print the filtered data to check
print("Step 2: Filtered data sample on US firms and years 1996-2015:\n", filtered_data.head(22))
print()

# To understand how many rows were dropped out
num_rows2 = len(filtered_data)
print("Number of rows in filtered data sample on US firms and years 1996-2015:", num_rows2)
print()
print()
```
```{python}
#| label: step3
#| echo: true
#| output: true

# Identify Year 0 to Year 6 for each firm within the filtered data
filtered_data['year_0'] = filtered_data.groupby('isin')['year_'].transform('min')  # this column will hold the according “Year 0” for each firm
filtered_data['relative_year'] = filtered_data['year_'] - filtered_data['year_0']  # this column now indicates how many years have passed since the firm’s “Year 0”
filtered_data = filtered_data[filtered_data['relative_year'] <= 6]  # filter the data to include only the first seven years (Year 0 to Year 6) for each firm

# Print the filtered data to check if it's empty or has data
print("Step 3: Filtered data sample after identifying the Year 0 and relative years:\n", filtered_data.head(22))
print()

# To understand how many rows were dropped out
num_rows3 = len(filtered_data)
print("Number of rows in filtered data sample after identifying the Year 0 and relative years:", num_rows3)
print()
print()
```

```{python}
#| label: step4
#| echo: true
#| output: true

# Calculate P/B Ratio for each firm in Year 0

year_0_data = filtered_data[filtered_data['relative_year'] == 0].copy()  # create a copy of filtered_data DataFrame to contain only data for Year 0
print("Step 4: Year 0 data:\n", year_0_data)
print()

year_0_data['bve'] = year_0_data['bve'].replace([0, np.inf, -np.inf], np.nan)  # replace zero and very small BVE values with NaN
# To verify: display rows where 'bve' is NaN
nan_bve_rows = year_0_data[year_0_data['bve'].isna()]
print("Rows where 'bve' has been replaced with NaN:\n", nan_bve_rows)
num_nan_bve_rows = len(nan_bve_rows)
print("Number of rows where 'bve' has been replaced with NaN:", num_nan_bve_rows)
print()

# Use P/B Ratio formula to calculate it for each firm in Year 0
year_0_data['p/b'] = year_0_data['mve'] / year_0_data['bve']

# Before removing NaN P/B ratios
num_rows_before = len(year_0_data)
# Remove rows with NaN P/B ratios
year_0_data = year_0_data.dropna(subset=['p/b'])
# After removing NaN P/B ratios
num_rows_after = len(year_0_data)

# Calculate the number of dropped observations
num_dropped = num_rows_before - num_rows_after
print("Number of observations dropped due to NaN P/B ratios:", num_dropped)  # checked: the number of dropped observations is equal to number of dropped BVE observations
print("Number of rows in Year 0 data after removing NaN P/B ratios:", num_rows_after)
print("Year 0 data after calculating P/B ratio and dropping NaN values:\n", year_0_data)  # the DataFrame now contains new column p/b ratio
print()
print()
```

```{python}
#| label: step5
#| echo: true
#| output: true

# Form P/B groups

# Sort firms based on their P/B ratios for Year 0 in descending order
num_groups = 20  # define the number of groups
year_0_data = year_0_data.sort_values('p/b', ascending=False).reset_index(drop=True)  # sort the year_0_data DataFrame in descending order based on the P/B ratios
print("Step 5: Year 0 data sorted in descending order based on the P/B ratios:\n", year_0_data)  # check the correct order
print()

# Assign the groups so that the highest P/B ratios are in group 1 and lowest in group 20
year_0_data['p/b_group'] = pd.qcut(year_0_data.index, num_groups, labels=range(1, num_groups + 1))
print("Year 0 data updated with P/B groups:\n", year_0_data)  # checked: p/b ratios grouping aligns with previous output

# Count the number of firms in each P/B group before merging
pb_group_counts_pre_merge = year_0_data.groupby('p/b_group', observed=True)['isin'].nunique()
print("Number of firms assigned to each P/B group before merging:\n", pb_group_counts_pre_merge)
print()

# Calculate median P/B value for each group
median_pb_values = year_0_data.groupby('p/b_group', observed=True)['p/b'].median()
# Display the median P/B values
print("Median P/B values for each group:\n", median_pb_values)
print()

# Merge the P/B groups back to the filtered data for Year 0
filtered_data = filtered_data.merge(year_0_data[['isin', 'p/b_group']], on='isin', how='left')
# Fill missing P/B groups for years other than Year 0 with the same group as Year 0
filtered_data['p/b_group'] = filtered_data.groupby('isin')['p/b_group'].ffill()

# Display the combined filtered data after merging P/B groups
print("Filtered data after merging P/B groups:\n", filtered_data.head(22))  # the DataFrame now contains new column p/b_group
# Count the number of firms in each P/B group after merging
pb_group_counts_post_merge = filtered_data.groupby('p/b_group', observed=True)['isin'].nunique()
print("Number of firms assigned to each P/B group after merging:\n", pb_group_counts_post_merge)  # checked: same number as pre merging
# Calculate the number of rows in the merged filtered data sample
num_rows_merged_filtered = len(filtered_data)
print("Number of rows in the merged filtered data sample:", num_rows_merged_filtered)  # checked: same number as pre merging
print()
```

```{python}
#| label: step6
#| echo: true
#| output: true

# Calculate residual income for each P/B group for Years 0 to 6, deflated by the book value of equity at the end of the year before Year 0 (= beginning of year 0)

cost_of_equity = 0.08  # assumption for cost of equity capital

# Create a copy of the merged data from Step 1 to preserve all years of observations
us_full_data = merged_data[merged_data['country'] == 'UNITED STATES'].copy()
print("Step 6: Merged data (Step 1) filtered to US only, all years preserved:\n", us_full_data.head(22))
print()

# Identify Year 0 for each firm, ensuring Year 0 is between 1996 and 2015
us_full_data['year_0'] = us_full_data.groupby('isin')['year_'].transform(lambda x: x[(x >= 1996) & (x <= 2015)].min())
us_full_data = us_full_data[us_full_data['year_0'].notna()]  # remove rows where year_0 is NaN
# Calculate the relative year and year_minus_1
us_full_data['relative_year'] = us_full_data['year_'] - us_full_data['year_0']
us_full_data['year_minus_1'] = us_full_data['year_0'] - 1

# Convert years to integer
us_full_data['year_0'] = us_full_data['year_0'].astype(int)
us_full_data['relative_year'] = us_full_data['relative_year'].astype(int)
us_full_data['year_minus_1'] = us_full_data['year_minus_1'].astype(int)
# Print the data to verify the steps
print("Data after identifying Year 0, relative year, and Year -1:\n", us_full_data.head(32))
print()

# Extract BVE for the year before Year 0 (relative_year = -1) for deflation
bve_deflation = us_full_data[us_full_data['relative_year'] == -1][['isin', 'bve']].rename(columns={'bve': 'bve_deflation'})
print(bve_deflation)
# Merge the bve_deflation values back to the filtered data
filtered_data = filtered_data.merge(bve_deflation, on='isin', how='left')
print("Filtered data with bve_deflation:\n",filtered_data)  # new column on bve_deflation
print()

# Calculate the BVE of the previous year for each year (for residual income calculation)
us_full_data['bve_prev'] = us_full_data.groupby('isin')['bve'].shift(1)
# Merge bve_prev back to filtered_data
filtered_data = filtered_data.merge(us_full_data[['isin', 'year_', 'bve_prev']], on=['isin', 'year_'], how='left')
print("Filtered data with bve_prev:\n",filtered_data)  # new column on bve_prev
print()

# Function to calculate non-deflated residual income
def calculate_residual_income(row):
    return row['ninc'] - (cost_of_equity * row['bve_prev'])
# Apply the function to calculate non-deflated residual income for each row
filtered_data['residual_income'] = filtered_data.apply(calculate_residual_income, axis=1) 
# Display the 'residual_income' values
print("Residual income values:\n", filtered_data)  # new column on residual_income
print()

# Function to deflate residual income
def deflate_residual_income(row):
    if pd.notna(row['bve_deflation']) and row['bve_deflation'] != 0:
        return row['residual_income'] / row['bve_deflation']
    else:
        return np.nan
# Apply the function to deflate residual income for each row
filtered_data['deflated_residual_income'] = filtered_data.apply(deflate_residual_income, axis=1)
print("Deflated residual income values:\n", filtered_data)  # new column on deflated_residual_income
print()
```

```{python}
#| label: step7
#| echo: true
#| output: true

# Output the replicated table

# Group by P/B group and relative year, then calculate the median deflated residual income for each group and year
median_residual_income = filtered_data.groupby(['p/b_group', 'relative_year'], observed=True)['deflated_residual_income'].median().unstack()

# Formatting the results (decimal places)
median_residual_income = median_residual_income.apply(lambda col: col.map(lambda x: f"{x:.3f}" if pd.notna(x) else "NaN"))
median_pb_values = median_pb_values.apply(lambda x: f"{x:.2f}")

# Add the median P/B values column to the output table
median_residual_income.insert(0, 'P/B', median_pb_values)

# Display the median residual income table
print("Median Residual Income by P/B Group and Relative Year:\n", median_residual_income)
print()
```

```{python}
#| label: step8
#| echo: true
#| output: true

# Step 8: Modification - exclude negative and very high P/B values

# Filter out extreme P/B values that are negative and higher than 7
filtered_year_0_data = year_0_data[(year_0_data['p/b'] > 0) & (year_0_data['p/b'] <= 7)]
print("Filtered Year 0 data after excluding negative and very high P/B values:\n", filtered_year_0_data)
print()

# Recalculate the P/B groups based on the filtered data
filtered_year_0_data = filtered_year_0_data.sort_values('p/b', ascending=False).reset_index(drop=True)
filtered_year_0_data['p/b_group'] = pd.qcut(filtered_year_0_data.index, num_groups, labels=range(1, num_groups + 1))
print("Filtered Year 0 data updated with P/B groups:\n", filtered_year_0_data)
print()

# Merge the P/B groups back to the filtered data for Year 0
filtered_data = filtered_data.drop(columns=['p/b_group'])
filtered_data = filtered_data.merge(filtered_year_0_data[['isin', 'p/b_group']], on='isin', how='left')
filtered_data['p/b_group'] = filtered_data.groupby('isin')['p/b_group'].ffill()

# Calculate median P/B values again
median_pb_values = filtered_year_0_data.groupby('p/b_group', observed=True)['p/b'].median()

# Group by P/B group and relative year, then calculate the median deflated residual income for each group and year
median_residual_income_filtered = filtered_data.groupby(['p/b_group', 'relative_year'], observed=True)['deflated_residual_income'].median().unstack()

# Formatting the results (decimal places)
median_residual_income_filtered = median_residual_income_filtered.apply(lambda col: col.map(lambda x: f"{x:.3f}" if pd.notna(x) else "NaN"))
median_pb_values = median_pb_values.apply(lambda x: f"{x:.2f}")

# Add the median P/B values column to the output table
median_residual_income_filtered.insert(0, 'P/B', median_pb_values)

# Display the median residual income table
print("Median Residual Income by P/B Group and Relative Year:\n", median_residual_income_filtered)
```

# Conclusion
This project demonstrates the effectiveness of using an open science and collaborative workflow for analyzing residual incomes and Price-to-Book ratios of U.S. public firms. By following a step-by-step approach and using the TRR 266 Template for Reproducible Empirical Accounting Research, I was able to replicate an empirical table from @Penman_2013 and provide insights into the trends of residual earnings across different P/B groups, similar to @Penman_2013.

The analysis revealed that high P/B groups initially exhibit higher residual earnings, which slowly decline over time, while low P/B groups consistently show negative residual earnings. Excluding extreme P/B values resulted in more stable and consistent residual earnings across groups and years.

In the future, this repository can be cloned or forked (if made public) to kickstart further projects on residual income analysis. Thanks for reading and enjoy!

\setcounter{table}{0}
\renewcommand{\thetable}{\arabic{table}}

# References {-}
\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\noindent
